---
title: "Data analysis for WP6"
author: "Patrizia Maier"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---


```{r setup, include=F, cache=F}

library(tidyverse)
library(gtsummary)
library(tidyverse)
library(janitor)
library(kableExtra)
# install.packages('tinytex')
# tinytex::install_tinytex() # latex for pdf file creation

```


```{r load_data, include=F}
in_file <- "WP6_data/WP6_individual_data.Rdata"
load(in_file)
rm(in_file)

data_individual <- data_individual %>% 
  mutate(Group=fct_rev(Group))

trial_file <- "WP6_data/WP6_trial_data.Rdata"
load(trial_file)
rm(trial_file)

```


## Sample demographics
Cross-sectional comparison between

* ALS (n = `r length(unique(data_individual$ID[data_individual$Group=="ALS"]))`) 
* Control (n = `r length(unique(data_individual$ID[data_individual$Group=="Ctrl"]))`)


```{r demo_all, echo=F, message=F, warning=F}

t1 <- data_individual %>% 
  select(Group, dfb_q1_sex, dfb_q2_age, dfb_q3_years_school_clean, dfb_q3_years_edu_total,
         dfb_q5_language_german, dfb_q6_handiness, dfb_q21_comp_expertise, dfb_q22_comp_freq, 
         sbsds_total_score) %>%  
  tbl_summary(by=Group, 
              label=list(dfb_q1_sex ~ "Gender", dfb_q2_age ~ "Age", 
                         dfb_q3_years_school_clean ~ "Years schooling",
                         dfb_q3_years_edu_total ~ "Years total education",
                         dfb_q5_language_german ~ "German native speaker", 
                         dfb_q6_handiness ~ "Handedness", 
                         dfb_q21_comp_expertise ~ "Self-rated computer expertise", 
                         dfb_q22_comp_freq ~ "Self-rated computer use frequency",
                         sbsds_total_score ~ "Self-rated spatial abilities (SBSDS)"),
              type=list(dfb_q3_years_school_clean ~ 'continuous',
                        dfb_q21_comp_expertise ~ 'continuous', 
                        dfb_q22_comp_freq  ~ 'continuous'),
              statistic=list(all_continuous() ~ "{median} (IQR: {p25}-{p75})",
                             all_categorical() ~ "{n} ({p}%)"),
              digits=list(all_continuous() ~ c(1, 2, 1, 1)),
              missing="no") %>% 
  add_p(test=list(all_continuous() ~ "wilcox.test", all_categorical() ~  "chisq.test")) %>% 
  modify_header(label = "**Variable**")

t1 %>% as_flex_table()

# t1 %>%
#   as_flex_table() %>%
#   flextable::save_as_docx(path="WP6_data/ALS_CTRL_Demographics.docx")

```

```{r demo_clinical, echo=F, message=F, warning=F}

t2 <- data_individual %>% 
  filter(Group=="ALS") %>% 
  select(MN, MNE_Untergruppe, ALS_Variante, `ALS-FRS-R`, `FRS-/Monat`,
         is_category, id_t1_months, is_t1_months) %>% 
  tbl_summary(label=list(MN ~ "Motor neuron involvement", 
                         MNE_Untergruppe ~ "MND subgroup", 
                         ALS_Variante ~ "ALS variant",
                         is_category ~ "Initial symptoms", 
                         id_t1_months ~ "Time from diagnosis (months)",
                         is_t1_months ~ "Time from initial symptoms (months)"),
              statistic=list(all_continuous() ~ "{median} (IQR: {p25}-{p75})",
                             all_categorical() ~ "{n} ({p}%)"),
              digits=list(all_continuous() ~ c(1, 1))) %>% 
  add_n() %>%
  modify_header(label = "**Variable**")

t2 %>% as_flex_table()

# t2 %>%
#   as_flex_table() %>%
#   flextable::save_as_docx(path="WP6_data/ALS_Clinical.docx")

```

## Results scoring (drawing) task

```{r results_scoring, echo=F, message=F, warning=F}

effect_size <- function(data, variable, by, ...) {
  rstatix::wilcox_effsize(data, as.formula(glue::glue("{variable} ~ {by}")))$effsize
}

t3 <- data_individual %>% 
  select(Group, Object_identity_manual_s, Object_location_SQRTCanOrg_s,
         Maze_reconstruction_manual_s) %>% 
  tbl_summary(by=Group,
              type=list(everything() ~ 'continuous'),
              label=list(Object_identity_manual_s ~ "Object identity", 
                         Object_location_SQRTCanOrg_s ~ "Object location", 
                         Maze_reconstruction_manual_s ~ "Maze reconstruction"),
              statistic=list(all_continuous() ~ "{median} ({p25}-{p75})"),
              digits=list(all_continuous() ~ c(2, 2)),
              missing="no") %>% 
    add_stat(fns=everything() ~ effect_size) %>%
  add_p(test=list(everything() ~ "wilcox.test"), 
        pvalue_fun=function(x) style_number(x, digits=3)) %>%  
  modify_header(list(label="**Post-navigation memory**", add_stat_1="**effect size**"))

t3 %>% as_flex_table()

# t3 %>%
#   as_flex_table() %>%
#   flextable::save_as_docx(path="WP6_data/ALS_CTRL_Scoring.docx")

```

## Results neuropsychology 

```{r results_np, echo=F, message=F, warning=F}

effect_size <- function(data, variable, by, ...) {
  rstatix::wilcox_effsize(data, as.formula(glue::glue("{variable} ~ {by}")))$effsize
}

t4 <- data_individual %>% 
  select(Group, ECAS_sub_executive, ECAS_sub_verbal_fluency, ECAS_sub_language,
         ECAS_sub_memory, ECAS_sub_spatial, 
         SPART_mean_I, SPART_mean_II, SPART_mean_all, 
         FIVE_P_productivity, 
         PTSOT_mean_dev) %>% 
  tbl_summary(by=Group,
              type=list(everything() ~ 'continuous'),
              label=list(ECAS_sub_executive ~ "ECAS executive",
                         ECAS_sub_verbal_fluency ~ "ECAS fluency",
                         ECAS_sub_language ~ "ECAS language",
                         ECAS_sub_memory ~ "ECAS memory", 
                         ECAS_sub_spatial ~ "ECAS visuospatial", 
                         SPART_mean_I ~ "SPART immediate",
                         SPART_mean_II ~ "SPART delayed",
                         SPART_mean_all ~ "SPART memory",
                         FIVE_P_productivity ~ "5PT fluency",
                         PTSOT_mean_dev ~ "PTSOT orientation"),
              statistic=list(everything() ~ "{median} ({p25}-{p75})"),
              digits=list(everything() ~ c(2, 2)),
              missing="no") %>% 
  add_stat(fns=everything() ~ effect_size) %>%
  add_p(test=list(everything() ~ "wilcox.test"), 
        pvalue_fun=function(x) style_number(x, digits=3)) %>% 
  modify_header(list(label="**Neuropsychology**", add_stat_1="**effect size**"))

t4 %>% as_flex_table()

# t4 %>%
#   as_flex_table() %>%
#   flextable::save_as_docx(path="WP6_data/ALS_CTRL_Neuropsychology.docx")

```

```{r results_np_ci1, echo=F, message=F, warning=F}

temp <- data_individual %>% 
  select(ID, Group, ends_with("below_cut"))

t5 <- temp %>% 
  select(-ID) %>% 
  tbl_summary(by=Group) %>% 
  add_n()

t5 %>% as_flex_table()

temp <- temp %>% 
  drop_na()

```

Participants below executive cut-off: `r temp$ID[temp$ECAS_sub_executive_below_cut]`

Participants below fluency cut-off: `r temp$ID[temp$ECAS_sub_verbal_fluency_below_cut]`

Participants below language cut-off: `r temp$ID[temp$ECAS_sub_language_below_cut]`

Participants below memory cut-off: `r temp$ID[temp$ECAS_sub_memory_below_cut]`

Participants below visuospatial cut-off: `r temp$ID[temp$ECAS_sub_spatial_below_cut]`


## Correlations
```{r results_correlation, echo=F, message=F, warning=F}

temp <- data_individual  %>% select(ID, Group, ECAS_sub_memory, dfb_q1_sex, dfb_q2_age, dfb_q3_years_edu_total,
                                    `ALS-FRS-R`, `FRS-/Monat`, is_t1_months, id_t1_months) %>%
  drop_na(ECAS_sub_memory) %>% filter(Group=="ALS")

ggplot(temp, aes(x=`ALS-FRS-R`, y=ECAS_sub_memory)) + geom_smooth(method="lm") + geom_point()
cor.test(temp$ECAS_sub_memory, temp$`ALS-FRS-R`, method="spearman")

ggplot(temp, aes(x=`FRS-/Monat`, y=ECAS_sub_memory)) + geom_smooth(method="lm") + geom_point()
cor.test(temp$ECAS_sub_memory, temp$`FRS-/Monat`, method="spearman")

ggplot(temp, aes(x=is_t1_months, y=ECAS_sub_memory)) + geom_smooth(method="lm") + geom_point()
cor.test(temp$ECAS_sub_memory, temp$is_t1_months, method="spearman")

ggplot(temp, aes(x=id_t1_months, y=ECAS_sub_memory)) + geom_smooth(method="lm") + geom_point()
cor.test(temp$ECAS_sub_memory, temp$id_t1_months, method="spearman")

ggplot(temp, aes(x=dfb_q2_age, y=ECAS_sub_memory)) + geom_smooth(method="lm") + geom_point()
cor.test(temp$ECAS_sub_memory, temp$dfb_q2_age, method="spearman")

ggplot(temp, aes(x=dfb_q3_years_edu_total, y=ECAS_sub_memory)) + geom_smooth(method="lm") + geom_point()
cor.test(temp$ECAS_sub_memory, temp$dfb_q3_years_edu_total, method="spearman")

m0 <- glm(ECAS_sub_memory ~ dfb_q1_sex + dfb_q2_age + dfb_q3_years_edu_total + 
            `ALS-FRS-R` + `FRS-/Monat` + is_t1_months, data=temp) 
summary(m0)
performance::r2(m0)

rm(m0)
```


## Regressions

```{r result_regression, include=F, message=F, warning=F }

sm_avg_condition <- trial_data %>%
  filter(feedback=="false", trial_condition %in% c(1,2,3), 
         trial_type!="probe_allo_startE") %>%
  group_by(id, trial_condition) %>%
  mutate(trial_condition=factor(case_when(trial_condition==1 ~ "allo", 
                                          trial_condition==2 ~ "ego", 
                                          TRUE ~ "mixed"))) %>% 
  summarize(avgSuccess=mean(success)) %>%
  pivot_wider(id_cols = c(id),
              names_prefix = "avgSuccess_", 
              names_from = trial_condition,
              values_from = c(avgSuccess))

sm_avg_total <- trial_data %>%
  filter(feedback=="false", trial_condition %in% c(1,2,3), 
         trial_type!="probe_allo_startE") %>%
  group_by(id) %>%
  summarize(avgSuccess_total=mean(success))


data_joint <- data_individual %>% 
  left_join(sm_avg_condition, by = c("ID"="id")) %>% 
  left_join(sm_avg_total, by = c("ID"="id"))

rm(sm_avg_condition, sm_avg_total)

```

## demographics and clinical markers

``` {r result_regression1, echo=F, message=F, warning=F }

m1 <- glm(avgSuccess_total ~ dfb_q1_sex + dfb_q2_age + dfb_q3_years_edu_total +
            MNE_Untergruppe + `ALS-FRS-R` + `FRS-/Monat` + is_t1_months, data=data_joint) 
#plot(m1)
summary(m1)
performance::r2(m1)

rm(m1)

```

## neuropsychology

``` {r result_regression2, echo=F, message=F, warning=F }
m2 <- glm(avgSuccess_total ~ ECAS_sub_executive + ECAS_sub_verbal_fluency + 
            ECAS_sub_spatial + ECAS_sub_memory + ECAS_sub_language +
            FIVE_P_productivity + SPART_mean_all + PTSOT_mean_dev, data=data_joint)
#plot(m2)
summary(m2)
performance::r2(m2)

rm(m2)

```

## task-related  

``` {r result_regression3, echo=F, message=F, warning=F }
m3 <- glm(avgSuccess_total ~ Object_identity_manual_s + 
            Object_location_SQRTCanOrg_s + 
            Maze_reconstruction_manual_s, data=data_joint)
#plot(m3)
summary(m3)
performance::r2(m3)

m3a <- glm(avgSuccess_ego ~ Object_identity_manual_s + 
             Object_location_SQRTCanOrg_s +
             Maze_reconstruction_manual_s, data=data_joint)
#plot(m3a)
summary(m3a)
performance::r2(m3a)

m3b <- glm(avgSuccess_allo ~ Object_identity_manual_s + 
             Object_location_SQRTCanOrg_s + 
             Maze_reconstruction_manual_s, data=data_joint)
#plot(m3b)
summary(m3b)
performance::r2(m3b)

rm(m3, m3a, m3b)

```

## Recruting 

```{r recruiting, echo=F, message=F, warning=F}

read_sheets <- function(filename, tibble = FALSE) {
  sheets <- readxl::excel_sheets(filename)
  x <- lapply(sheets, function(X) readxl::read_excel(filename, sheet = X, skip=2))
  if(!tibble) x <- lapply(x, as.data.frame)
  names(x) <- sheets
  x
}

filename1 <- "WP6_data/Rekrutierung_210820.xlsx"
mysheets <- read_sheets(filename1)
names_invite <- do.call(rbind.data.frame, mysheets) %>% 
  select(-c(Alter, Adresse, `Termin Tag`, `Termin Uhrzeit`, Rückmeldung, Notizen)) %>% 
  filter(`Brief verschickt` != "nein") %>% 
  select(-c(`Brief verschickt`)) %>% 
  unique() %>% 
  unite(Vorname, Nachname, col="Name", sep=" ") %>% 
  arrange(Name) 

filename2 <- "WP6_data/WP6_ID_List_6200_6399.xlsx"
names_participate <- readxl::read_excel(filename2, skip=1, n_max=52, trim_ws = T) %>% 
  select(Name) %>% 
  arrange(Name)

rm(filename1, filename2, read_sheets)
```

### People that were recruited via mail (with help of ALS-Ambulanz)
N of letters = `r length(names_invite$Name) `

N of respondents = `r length(names_invite$Name[names_invite$Name %in% names_participate$Name]) `
(`r round(length(names_invite$Name[names_invite$Name %in% names_participate$Name])/length(names_invite$Name)*100, digits=2)`)

### People that were recruited elsewhere (e.g. from the Neurology ward)
N of respondents = `r length(names_participate$Name[!names_participate$Name %in% names_invite$Name]) `
